Decline in U.S Birthrates is Indicative of Socio-economic Changes

### 2022-02-27 Output/paper 2.Rmd
adding changes

---
better economic-industrial society to expensive for children, industrial has low birth rates
adoption of more freedom for women?
abortion? restricting of abortion could lead to higher birth rats, but to the detriment of choice

Anyone born between 1978 and 1996 (ages 23 to 38 in 2019) is considered a Millennial, and anyone born from 1997 onward is part of a new

why is hispanic decline so large
harmful ideologies: white people decreasing birht and population, hispanics increasing births and popualtion., hire more immigrants

implications for teenaged mom is a good thing that the borth rates are decreasing?

Check how teen pregnancy  and female enollment has increased.

Its good black teen birth rate decreased (table 1), how can we increase this decrease?

does birth control reduce usage for young mothers?
we dont expect income ratio to affect as these are young moms anyway.
religious change, specially in bible belt?

possible bias: To what extent is random allocation ethical and fair? Some argue that shortages mean it is reasonable to randomly allocate, but that may depend on how linear the benefits are. It may also be difficult to establish definitions, and the power imbalance between those making these decisions and those being treated should be considered

@@

A nation's birthrate for a particular year is the total number of live births per 1000 people for that year. In order to measure the birth rates and therefore fertility levels, as well as forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors. The total population of the country is the total number of citizens and other residents in the country, regardless of legal status. It is used in many types of planning, such as planning and decisions on public services, health care, education, government funding, transportation, and politics. The total population is obtained through a population census, provided by the CDC Surveillance, Epidemiolody, and End results program (CDC SEER). This information also breaks down the population by age, race, and socio-economic factors. In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographic of the mother. We are also interested in the relationship between  changes in economic, social and cultural aspects of the United states and the changes in these birth rates.

We obtained the dataset from the American Economic Association [@citewebsite]. We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr], `janitor` [@citejanitor], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegrid] and `kableExtra` [@citekable]. We first created histograms and scatterplots which showed the most important variables. We then created line plots and a map which showed the trend in birth rates over the years, for different age groups, races, and states. We then created tables which showed the relative contributions of our target demographics on the change in birth rates and the differences in their populations, as well as showed the change in values for possible economic and societal factors that may be correlated to birth rates.

@@

While this census is widely used to forecast population growth and make important decisions on the future of the country, biases and problems in data collection could paint an incorrect view of birth rates and the total population. The declining birth rates also brings to light the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry]. This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford kids without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their systematic discrimination in society. The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being "irresponsible" with their bodies. This is because the data only collects the statistics but does not provide meaning, causes, and explanations behind these statistics [@Gold2001].  There are also concerns about biases in data collection of data, political impacts of the data as well as the future state of population growth in the US and its consequences.

@@

The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite]. The data used by the paper was compiled by various sources such as the NCHS [@citenchs] CDC SEER [@citecdcseer], NBER Natality Database, and the Current Population Survey
Annual Social and Economic Supplement (CPS ASEC), the New York Federal Reserve, and the Integrated Public Use Microdata Series (IPUMS) (cite all). The paper and datasets were last accessed on February 16th, 2022. The first two sources collect and estimate data via national census from US citizens and residents by county or state over the years 1968 to 2019. The birth rate data was gotten from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, which was randomly sampled [@citenchs]. The national population data from the census counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer]. The datasets from the other sources compiled by the paper authors typically include data by year and by state for several demographics. These datasets were gotten from surveys or website data, so non-response was often recorded as 0 or as `NA`. The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity. Meanwhile, categorical data mainly appeared in the form of state.

@@

We included a dataset if they were considered important to answering the question behind birthrates, or revealed important information. We included the `race` dataset as it gave the main data of analysis  in the form of birth rates for different age groups and races. We included the `birth` dataset as it included the total birthrate for all women. We also included the `numbirth` dataset as it had information on birthrates for all the age groups combined. We used the `birth1` dataset as well as it had data on birth rates for younger age groups, as well as total population data and demographic data for each race and age combination. `birth2` had total population data for each age group and race-age combination. We used`policy`, `rent`, `student_debt`, `religion`, and `childcare` as they provided information on the changes in unemployment, rental cost, student debt, importance of religion and rise in child costs throughout the US, all of which are expected to influence birth rates to varying degrees.

We created several of our own datasets by modifying and joining variables.  These datasets include `birthhist`, which was gotten from the `birth1` dataset by adding all the educational status for each age-race combination to get the total birth race for that comnination, and then compiling the values by state and year. We also got the `sumpop` variable by adding the total population for each state in a year.  We then obtained `mapdiff`, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, `table1`, which was used to show how our target demographics contributed to birth rate decline, and `explanatory`, which was used to display the changes in possible explanatory variables over a period of time. Most final datasets include either 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).

We modified many variables for the report. We first modified the `birth1` dataset to add variables for the number of births for each race and age group combination, as they were unavailable in the original dataset. We did this by adding up all educational statuses for each race and age group combination, as each person whose birth rates was recorded also had their educational status recorded.  The final variables were named `numbirthraceage`, i.e `numbirthwhite15` for births for the white 15-19 population. 
We then modified the `race` dataset to change the columns of 15-19 and 20-24 age to rows within the column of `group`. We also modified the `birthhist` dataset's columns to be rows within the column of `racegroup`, then multiplied by 10000 and divided by the total female population to get the birthrates. These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.

For the `mapdiff` dataset, we obtained the value of `decline`, which showed the change in birth rates by state, by first creating two datasets that got the change in number of births and population (from the `birth1` and `birth2` datasets respectively) for the 20-34 age group from 1999-2000 and 2018- 2019 by state. The values for the 15-19 and 20-24 ages groups were unavailable for those datasets, so we needed to use the next best age group. From those two new datasets, we changed the `year` column so that it only stated 2000 or 2019. We then joined those two datasets together and divided the number of births by the population to get the birth rate. We then created two additional datasets that contained only the values of 2000 and only the values of 2019, and collected the mean of each value. To get the final dataset of `mapdiff`, we joined those two datasets by state, and got the `decline` variable by subtracting the average birthrate of 2019 from that of 2000.

For the first table, we got the `table` dataset by grouping the `numbirth` dataset by year, beyond 2000, and then summing up the entire female population by state, which was stored as `sumbirth`. We then created the `ageracepop` dataset using the same methods but now summing up the different age-race combination populations. We named these variables for each combination `sumraceage`, e.g `sumblack20`. We filtered the `birth` dataset between 2000 and 2020. We then joined these three datasets, along with the previous `birthhist`, to create `fulltable`. From `fulltable` we got the `share_of_pop` dataset with variables that got the percentage of population of a demographic by dividing the `sumraceage` variable by the `sumpop` variable. From `fulltable` we also got `birthrates` which got the relative birthrates for each race-age combination, by multiplying the number of births for each race-age combination by 1000 and them dividing them by the total female population. These variables were named `raceagebrate`, e.g `b20brate`. We then got the `contribution` variable by subtracting 2019 values from the 2000 values and dividing that by the `brate_all` value for 2000 and multiplying by 200. We then got `percentpop` by multiplying the `share_of_pop` dataset's variables by 100.

For the second table, we got `avg_grossrent` by grouping the `student_debt` variable by year and getting the mean value of all the states. We got `meanchildcare` and `meandebt` using the same methods from the `childcare` and `studentdebt` datasets respectively. For `meandebt`, we multiplied it by 10,000 as that is how it was shown in the original source. We got the `meanpercentimportant` dataset by adding the values of `very important` and `somewhat important` from the `religion` dataset and dividing it by the previous two values plus `not_too_important` and `not_at_all_important`, and then getting the mean of all the states. Note that for these variables we needed to use 2004 as 2000 was not available.

```{r global_options, include=FALSE}

knitr::opts_chunk$set(fig.pos = 'H')
library("kableExtra")
# install.packages("usmap")
library("tidyverse")
library("haven")
library("janitor")
library("dplyr")
library("usmap")
library("gridExtra")
```

```{r,include = FALSE}
birth = read_csv(here::here("Inputs/Data/birth.csv"))
numbirth = read_csv(here::here("Inputs/Data/numbirth.csv"))
student_debt = read_csv(here::here("Inputs/Data/student_debt.csv"))
rent = read_csv(here::here("Inputs/Data/rent.csv"))
religion = read_csv(here::here("Inputs/Data/religion.csv"))
childcare = read_csv(here::here("Inputs/Data/childcare.csv"))
birth2 = read_csv(here::here("Inputs/Data/birth2.csv"))
birth1 = read_csv(here::here("Inputs/Data/birth1.csv"))
race = read_csv(here::here("Inputs/Data/race.csv"))
policy = read_csv(here::here("Inputs/Data/policy.csv"))
```
## View of the Data

We are interested in the trend of birth rates for women ages 15-19 and 20-24 over a 20 year period. This shows the decline in popularity of young motherhood and thus the increase of the average birth age. This has ongoing implications for public social and economic policy.  (Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of number of births for women in the two age groups for 20 years, sectioned by race.

We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant years being between 60 to 190 thousand. The data seems evenly skewed and unimodal. We see that teenage births tend to be less popular, as the data seems slightly right skewed. This also means that the mean births will be below the median births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012]. We also see that black teenage mothers are less common, as they have the lowest births for most years. This is most likely because they are a much smaller segment of the population than the others. However we see that black mothers still do manage to have a moderate amount of births, indicating they are disproportiantely appearing as teenage mothers. Hispanic teenage mothers are more middle ground, having between 60 to 100 thousand births per year. White teenage mothers are the most common, due to them being the largest in population. They tend to have most of their distribution in the high areas towards the higher numbers.

For the 20-24 age group, there seems to be a slight bimodal distribution, with many births being on two extremes. Due to this it is difficult to determine whether the mean or the median of births will be larger.  We see that there are far more women giving birth at this age group than in the previous age group. We see that black mothers are overwhelmingly on the lower end of births, which brings into question why teenage black mothers had more prevalent births compared to their elder counterparts. Hispanic mothers also tend to be at the lower end of the scale, but are more spread out. White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.

We see that births for black and hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while white mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in black and hispanic women may be more common.

In contrast, in (Figure \@ref(fig:figthree)) the distribution for the total population of women for the past 20 years seems uniformal, with almost all years having their own value. This is because the population of a nation either increases or decreases, rarely staying the same. Due to this we can say that the mean and median population are approximately the same. However there are some years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year. From the scatterplot we see that the total female population has only been increasing in recent years, with a seemingly exponential increase. This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birthrates. The two dips in the plot correspond to the previous histogram of years with similar value.

```{r, include=FALSE}
birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519

birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519

birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519

birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024

birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024

birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024

birthhist = birth1 %>%
  filter(year >= 2000) %>%
  group_by(year) %>% 
  select(year,pop1544, numbirthwhite15,numbirthwhite20,numbirthhisp15,numbirthhisp20,numbirthblack15,numbirthblack20) %>%
  summarize(sumpop = sum(pop1544),white15 = sum(numbirthwhite15), white20 = sum(numbirthwhite20), black15 = sum(numbirthblack15), black20= sum(numbirthblack20), hisp15 = sum(numbirthhisp15), hisp20=sum(numbirthhisp20))

@@

```{r figone, echo = FALSE, fig.cap = "Distribution of Number of Births for the 15 to 19 age group from 2000 to 2020"}
birthhist %>%
  pivot_longer(cols = c("white15", "black15","hisp15"),
               names_to = "Race",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Number of births") +
  ylab("Number of Years")

```
```{r figtwo, echo = FALSE, fig.cap = "Distribution of the number of births from the 20 to 24 age group from 2000 to 2020" }
birthhist %>%
  pivot_longer(cols = c("white20", "black20","hisp20"),
               names_to = "Race",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Number of Births") +
  ylab("Number of Years")

@@

```{r figthree, echo = FALSE, fig.cap = "Distribution of the total number of births and total female population from 2000 to 2020"}
par(mfrow=c(2,2))
pop_plot = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(population = sum(pop)) %>%
  ggplot(data =., aes(x=population)) +
  geom_histogram(color="black", fill="white",bins=20)+ 
  theme_minimal()+
  xlab("Total Female Population") +
  ylab("Number of Years")

pop_plot1 = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(population = sum(pop)) %>%
  ggplot(aes(x = year, y =population)) +
  geom_point()+ 
  theme_minimal()+
  xlab("Year") +
  ylab("Total Female Population")

grid.arrange(pop_plot,pop_plot1, nrow =2)
```

@@

race %>%
  select(year,brate_1519,brate_2024) %>%
  filter(year >= 2000) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = group)) +
  geom_line() +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000,2020,by = 2))+
  scale_y_continuous(breaks = seq(0,140,by = 20))+
  labs(x = "Year",
       y = "Birth Rate",
       color = "Age group") +
  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24")) +
  theme(legend.position = "bottom")
```

(Figure \@ref(fig:figfive)) shows the trend of births for different races of women from the ages of 15 to 24 over the past 20 years. We see that the birthrate for Hispanic mothers significantly decreased, going from about 63 births per 1000 population to just about 40 births for 1000 population. This can either be because the population of hispanic women has increased due to immigration, or that the number of births has decreased. The birth rates for black women has also halved, going from 50 births per 1000 to just 25 births per 1000. The white birth rates however has taken the largest decrease, going from about 125 births to about 65 births through the 20 years. We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation z they have converged more closely due to the decrease in birth rates for all races. This could be due to economic factors directly related to race, such as inuequality in the healthcare system [@Davis2018] . As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].

>>>>

---

@@

A nation's birthrate for a particular year is the total number of live births per 1000 people for that year. In order to measure the birth rates and fertility levels to forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors. The total population of the country is the total number of citizens and other residents in the country, regardless of legal status. It is used in many types of planning, such as planning and decisions on public services, health care, education, government funding, transportation, and politics. The total population is obtained through a population census, provided by the CDC Surveillance, Epidemiolody, and End results program (CDC SEER). This information also breaks down the population by age, race, and socio-economic factors. In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographics of the mother. We are also interested in the relationship between the change in economic, social and cultural aspects of the United states and the change in these birth rates.

We obtained the dataset from the American Economic Association [@citewebsite]. We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr], `janitor` [@citejanitor], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegrid] and `kableExtra` [@citekable]. We first created histograms and scatter plots which showed the most important variables. We then created line plots and a map which showed the trend in birth rates over the years, for different age groups, races, and states. We then created tables which showed the relative contributions of our target demographics on the change in birth rates and the differences in their populations, as well as showed the change in values for possible economic and societal factors that may be correlated to birth rates.

@@

While this census is widely used to forecast population growth and make important decisions on the future of the country, biases and problems in data collection could paint an incorrect view of birth rates and the total population. The declining birth rates also brings to light the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry]. This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford kids without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their discrimination in society. The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being irresponsible with their bodies. This is because the data only collects the statistics but does not provide meaning, causes, and explanations behind these statistics [@Gold2001].  There are also concerns about biases in data collection of data, political impacts of the data as well as the future state of population growth in the US and its consequences.

@@

The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite]. The data used by the paper was compiled by various sources such as the NCHS Natality Database [@citenchs], CDC SEER [@citecdcseer],the Current Population Survey Annual Social and Economic Supplement (CPS ASEC) , the New York Federal Reserve [@citefed], and the Integrated Public Use Microdata Series (IPUMS) [@citeipums]. The paper and datasets were last accessed on February 16th, 2022. The first two sources collect and estimate data via national census from US citizens and residents by county or state over the years 1968 to 2019. The birth rate data was gotten from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, and this was randomly sampled [@citenchs]. The national population data comes from the census, a long running survey funded by the government released on a predetermined cycle, and counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer]. The datasets from the other sources compiled by the paper authors typically include data by year and by state for several demographics. These datasets were gotten from surveys so not everyone responded, thus the data was often recorded as 0, `NA`, or estimated.The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity. Meanwhile, categorical data mainly appeared in the form of state.

@@

We included a dataset if they were considered important to answering the question behind birthrates, or revealed important information. We included the datasets that had information on race, as this was needed to compare birth rates for different races. We also had datasets on total number of births and/or birth rates, both of which were used to analyze the decline in births. We had total population datasets, used to get the birth rates and our demographic's share of the population. We had datasets on birth rates for different age groups, states, and years, all of which were used to analyze the trends in birth rates for different groups.  We also included datasets that included information on government policy, rent costs, student debt, religious beliefs, and childcare costs. These datasets were used to signify a possible relationship between these variables and birth rates.

We created several of our own datasets by modifying and joining variables. One dataset was gotten by adding all the educational statuses for each age-race combination from another dataset to get the total birth rate for each combination, and then compiling the values by state and year. We also got a variable for total population by adding the total population for each state in a year.  We obtained the final dataset used to create the map of birth rate decline by combining various other datasets by year, and subtracting the birth rates of the later years by that of the earlier years. The final dataset that showed how our demographics of interest contributed to total birth rate decline was created by joining several datasets all which compared important information, then dividing the columns of our demographic by the column of total birth rates to get the relative contributions. The final dataset used to display the change in explanatory variables over time was gotten by combining the datasets, each of which explained the changes in a single variable. Most final datasets include either 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).

We modified many variables for the report. We created variables for total birth rates of each race-age combination by adding up all educational statuses for each race and age group combination and dividing by the total population, as each person whose birth rates was recorded also had their educational status recorded. We also modified age group variables by changing the columns of 15-19 and 20-24 age to rows within the column of `group`. We did the same for race variables, modifying the columns to be rows within one group, and then multiplying the values column by 10,000 and dividing by the total female population to get the birth rates for each race.These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics. We obtained a variable which got the decline in birth rates from 2000 to 2019 by state, by creating dividing the number of births by population to get birth rates, grouping them by year and state, and then subtracting the values of the later year from the values of the earlier year.

We got a variable for the total number of births by adding up all the births by state and then grouping them by year. We got a variable that stated the total population of each race-age combination by similar methods. we obtained a variable got the percentage of population of a demographic by dividing the sum of that demographic by the total population and multiplying by 100. We had a variable that got contribution to total birth rate decline by subtracting 2019 values from the 2000 values and dividing that by the total birthrate value for 2000 and multiplying by 200. We got the variables for all our explanatory factors by grouping them by year and getting the mean values from all the states. For the average debt, we multiplied by 10,000 as that is how it was shown in the original source. We got the variable on the importance of religion by dividing the number of people who felt religion was important over the total number surveys.Note that for these variables we needed to use 2004 as 2000 was not available.

```{r global_options, include=FALSE}
# laoding in relevant libraries
knitr::opts_chunk$set(fig.pos = 'H')
library("kableExtra")
# install.packages("usmap")
library("tidyverse")
library("haven")
library("janitor")
library("dplyr")
library("usmap")
library("gridExtra")
```

```{r,include = FALSE}
#loading in relevant datasets
birth = read_csv(here::here("Inputs/Data/birth.csv"))
numbirth = read_csv(here::here("Inputs/Data/numbirth.csv"))
student_debt = read_csv(here::here("Inputs/Data/student_debt.csv"))
rent = read_csv(here::here("Inputs/Data/rent.csv"))
religion = read_csv(here::here("Inputs/Data/religion.csv"))
childcare = read_csv(here::here("Inputs/Data/childcare.csv"))
birth2 = read_csv(here::here("Inputs/Data/birth2.csv"))
birth1 = read_csv(here::here("Inputs/Data/birth1.csv"))
race = read_csv(here::here("Inputs/Data/race.csv"))
policy = read_csv(here::here("Inputs/Data/policy.csv"))
```

## View of the Data

We are interested in the trend of birth rates for women ages 15-19 and 20-24 over a 20 year period. The data shows the decline in popularity of young motherhood and thus the increase of the average birth age. This has ongoing implications for public social and economic policy.  (Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of number of births for women in the two age groups for 20 years, sectioned by race.

We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant years being between 60 to 190 thousand. The data seems evenly skewed and unimodal. We see that teenage births tend to be less popular, as the data seems slightly right skewed. This also means that the average number of births will be below the median number of births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012]. We also see that black teenage mothers are less common, as they have the lowest births for most years. This is most likely because they are a much smaller segment of the population than the others. However we see that black mothers still do manage to have a moderate amount of births, indicating they are disproportionately appearing as teenage mothers. Hispanic teenage mothers are more middle ground, having between 60 to 100 thousand births per year. White teenage mothers are the most common, due to them being the largest in population. They tend to have most of their distribution towards the higher numbers.

For the 20-24 age group, there seems to be a slight bimodal distribution, with many births being on two extremes. Due to this it is difficult to determine whether the mean or the median of births will be larger.  We see that there are far more women giving birth at this age group than in the previous age group. We see that black mothers are overwhelmingly on the lower end of births, which brings into question why teenage black mothers had more prevalent births compared to their elder counterparts. Hispanic mothers also tend to be at the lower end of the scale, but are more spread out. White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.We see that births for black and hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while white mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in black and hispanic women may be more common.

In contrast, in (Figure \@ref(fig:figthree)) the distribution for the total population of women for the past 20 years seems uniformal, with almost all years having their own value. This is because the population of a nation either increases or decreases, rarely staying the same. Due to this we can say that the mean and median population are approximately the same. However there are some years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year. From the scatterplot we see that the total female population has only been increasing in recent years, with a seemingly exponential increase. This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birthrates. However, the two dips in the scatterplot correspond to the previous histogram of which had years with similar values.

```{r, include=FALSE}
# getting total number of births for white people from 15 to 19 years old
birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519

# getting total number of births for black people from 15 to 19 years old  
birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519

# getting total number of births for hispanics from 15 to 19 years old  
birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519

# getting total number of births for white people from 20 to 24 years old
birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024

# getting total number of births for black people from 20 to 24 years old
birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024

# getting total number of births for hispanics from 20 to 24 years old  
birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024

# gets the total number of births by year, compiled nationally
birthhist = birth1 %>%
  filter(year >= 2000) %>%
  group_by(year) %>% 
  select(year,pop1544, numbirthwhite15,numbirthwhite20,numbirthhisp15,numbirthhisp20,numbirthblack15,numbirthblack20) %>%
  summarize(sumpop = sum(pop1544),white15 = sum(numbirthwhite15), white20 = sum(numbirthwhite20), black15 = sum(numbirthblack15), black20= sum(numbirthblack20), hisp15 = sum(numbirthhisp15), hisp20=sum(numbirthhisp20))

@@

```{r figone, echo = FALSE, fig.cap = "Distribution of Number of Births for the 15 to 19 age group from 2000 to 2020", message = FALSE}

# converts the columns to rows and creates the histogram
birthhist %>%
  pivot_longer(cols = c("white15", "black15","hisp15"),
               names_to = "Race",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Number of births") +
  ylab("Number of Years")

```
```{r figtwo, echo = FALSE, fig.cap = "Distribution of the number of births from the 20 to 24 age group from 2000 to 2020", message = FALSE}

# converts the columns to rows and creates the histogram

birthhist %>%
  pivot_longer(cols = c("white20", "black20","hisp20"),
               names_to = "Race",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Number of Births") +
  ylab("Number of Years")

@@

```{r figthree, echo = FALSE, fig.cap = "Distribution of the total number of births and total female population from 2000 to 2020"}

# groups the total female population by year and compiles nationally, then creates the histogram
pop_plot = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(population = sum(pop)) %>%
  ggplot(data =., aes(x=population)) +
  geom_histogram(color="black", fill="white",bins=20)+ 
  theme_minimal()+
  xlab("Total Female Population") +
  ylab("Number of Years")

# groups the total female population by year and compiles nationally, then creates the scatterplot  
pop_plot1 = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(population = sum(pop)) %>%
  ggplot(aes(x = year, y =population)) +
  geom_point()+ 
  theme_minimal()+
  xlab("Year") +
  ylab("Total Female Population")

# putting two figures into one figure space
grid.arrange(pop_plot,pop_plot1, nrow =2)
```

@@

# converting the age column to rows, then creates a line graphing the the trend of the two age groups
race %>%
  select(year,brate_1519,brate_2024) %>%
  filter(year >= 2000) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = group)) +
  geom_line() +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000,2020,by = 2))+ # labeling by 2
  scale_y_continuous(breaks = seq(0,140,by = 20))+ # labeling by 20
  labs(x = "Year",
       y = "Birth Rate",
       color = "Age group") +
  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24")) +
  theme(legend.position = "bottom")
```

(Figure \@ref(fig:figfive)) shows the trend of births for different races of women from the ages of 15 to 24 over the past 20 years. We see that the birthrate for Hispanic mothers significantly decreased, going from about 63 births per 1000 population to just about 40 births for 1000 population. This can either be because the population of hispanic women has increased due to immigration, or that the number of births has decreased. The birth rates for black women has also halved, going from 50 births per 1000 to just 25 births per 1000. The white birth rates however has taken the largest decrease, going from about 125 births to about 65 births through the 20 years. We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation Z they have converged more closely due to the decrease in birth rates for all races. This could be due to economic factors directly related to race, such as inequality in the healthcare system [@Davis2018] . As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].

```{r figfive, echo = FALSE, fig.cap = "Trend in Birth rates by Race from 2000 to 2020"}

# gets the birth rate for each race by adding the number of births from the 15 to 19 and 20 to 24 age group for each race, then multiplying by 10,000 and dividing by the female population
